### 2024-04-04 Chapters/Chapter2.tex
Uppfært eftir ábendingar ÓPP

Veðurstofan provided recordings from weather stations all around Iceland. These automatic weather stations (AWS) are at a height of 10 meters above ground level. The information that is provided by these AWS is presented in two different type of documents, hourly and 10 minute documents. The hourly documents are summations of the 10 minute documents, with the exception that any errors (so called nails) still in the 10 minute documents should have been removed from the hourly documents. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another document), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction. These features for each of the stations go back at least two decades, but do not have the same start date. This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.

@@

Each request is made by going through all instances of the Veðurstofu data were able to fulfill the requirements (of mean wind speed reaching \averageWindSpeedLimit m/s over a 10 minute interval). For each such observation in the Veðurstofa data, two API calls excluding redundancy are needed. One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly one of the three hour interval times. That is if the observation was at 13:00, we would need Carra data for noon and 15:00. Date and times were generated automatically from the AWS and requested by calling the client. Using these datapoints, interpolation was used to get an estimation for the point of the given weather station. The Carra data contains several types of layers. These are single levels, model levels, height levels, pressure levels. The data for this observation was downloaded from height levels. That is, data was requested at heights of 15, 150, 250 and 500 meters above ground. For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature. Each of these features needed to be interpolated to create data for model to be trained on.

>>>>

Veðurstofan provided recordings from 412 weather stations all around Iceland. The location of these weather stations can be seen in Figure \ref{fig:aws_map}. These automatic weather stations (AWS) are at a height of 10 meters above ground level. The information that is provided by these AWS is presented in two different type of documents, hourly and 10 minute documents. The hourly documents are summations of the 10 minute documents, with the exception that any errors (so called nails) still in the 10 minute documents should have been removed from the hourly documents. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another document), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction. These features for each of the stations go back at least two decades, but do not have the same start date. This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.

\begin{figure}[h]
    \caption{Locations of Automatic Weather Stations in Iceland}
    \label{fig:aws_map}
    \includegraphics[scale = 0.75]{Figures/weather_stations.png}
\end{figure}

@@

Each request is made by going through all instances of the Veðurstofu data were able to fulfill the requirements (of mean wind speed reaching \averageWindSpeedLimit m/s over a 10 minute interval). For each such observation in the Veðurstofa data, two API calls excluding redundancy are needed. One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly one of the three hour interval times. That is if the observation was at 13:00, Carra data for noon and 15:00 would have to be queried. Date and times were generated automatically from the AWS and requested by calling the client. Using these datapoints, interpolation was used to get an estimation for the point of the given weather station. The Carra data contains several types of layers. These are single levels, model levels, height levels, pressure levels. The data for this observation was downloaded from height levels. That is, data was requested at heights of 15, 150, 250 and 500 meters above ground. For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature. Each of these features needed to be interpolated to create data for model to be trained on.

### 2024-04-04 Chapters/Chapter3.tex
Uppfært eftir ábendingar ÓPP

The procedure of combining these sources was as follows. The measured data from the AWS is filtered by using a limit on the average wind speed. The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}). Even so being able to predict the gust factor is more important for higher average wind speed as there are the highest wind gusts. After this stripped dataset over every AWS has been created it is used to query the Carra data by using their API. The Carra API needs to be queried for given hours, days, months, years and a given area. That is, if queried for a given hour, it returns that hour for every day that is queried. Similarly if queried for a given day, it returns that day for every month. In light of these restraints, it was decided to query month by month. Querying only the days needed (both the days included in the 10 minute measurements and the days needed for bridging if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00 PM) and then take these values and bridge for the 10 minute values. After querying and downloading the data for the height levels and variables requested, bridging is done for the points of interest and values stored in a pandas dataframe. After this is done the downloaded data is discarded and the next month is queried. This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).

The elevation data comes in a GeoTif file that covers Iceland. It is a rectangular grid of resolution 20 meters. For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved. For each point retrieved bridging needs to be done. This is done in a similar manner to the interpolation of the Carra data. Weights are assigned to each of the four points bounding the point of interest. These weights are then used to calculate the weighted average that represents the bridged value. This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.

@@

Once data has been retrieved for all three sources and processed, including bridging values, it needs to be made ready to use by the model, for both training, validation and test. Starting with a dataframe that contains measured information from AWS. This includes the average wind, the wind gust, wind direction along with the station number and coordinates. When selecting the Carra data certain height levels are chosen. That is at which heights above ground the reanalysis parameters are requested. These present as separate lines in the Carra dataframe. That is information for one observation is represented in as many lines as height levels requested in the reanalysis data. These rows need to be combined on the position (the weather station) and time so that each row contains all reanalysis information for a given measurement of a weather station. When this is done it is possible to combine the AWS Veðurstofu data and Carra reanalysis data on the location and time columns. The last data source is the elevation. A couple of different sections of land around the weather stations have been looked at. A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point. In any case the points, that represent these sections, were selected like as shown in code Listing (\ref{code:sectorElevation}). That is a range of angles are defined based on the wind direction d at some distance from the given point. This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.

>>>>

\begin{figure}[h]
    \caption{A flow chart showing how data sources were combined}
    \label{fig:data_preprocessing_flow_chart}
    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
\end{figure}

The procedure of combining these sources was as follows and can be seen in Figure (\ref{fig:data_preprocessing_flow_chart}). The measured data from the AWS is filtered by using a limit on the average wind speed. The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}). Even so being able to predict the gust factor is more important for higher average wind speed as there are the highest wind gusts. After this stripped dataset over every AWS has been created it is used to query the Carra data by using their API. The Carra API needs to be queried for given hours, days, months, years and a given area. That is, if queried for a given hour, it returns that hour for every day that is queried. Similarly if queried for a given day, it returns that day for every month. In light of these restraints, it was decided to query month by month. Querying only the days needed (both the days included in the 10 minute measurements and the days needed for interpolation if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00 PM) and then take these values and interpolate for the 10 minute values. After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe. After this the downloaded data is discarded and the next month is queried. This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).

The elevation data comes in a GeoTif file that covers Iceland. It is a rectangular grid of resolution 20 meters. For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved. For each point retrieved interpolation needs to be done. This is done in a similar manner to the interpolation of the Carra data. Weights are assigned to each of the four points bounding the point of interest. These weights are then used to calculate the weighted average that represents the interpolated value. This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.

@@

Once data has been retrieved for all three sources and processed, including interpolating values, it needs to be made ready to use by the model, for both training, validation and test. Starting with a dataframe that contains measured information from AWS. This includes the average wind, the wind gust, wind direction along with the station number and coordinates. When selecting the Carra data certain height levels are chosen. That is at which heights above ground the reanalysis parameters are requested. These present as separate lines in the Carra dataframe. That is information for one observation is represented in as many lines as height levels requested in the reanalysis data. These rows need to be combined on the position (the weather station) and time so that each row contains all reanalysis information for a given measurement of a weather station. When this is done it is possible to combine the AWS Veðurstofu data and Carra reanalysis data on the location and time columns. The last data source is the elevation. A couple of different sections of land around the weather stations have been looked at. A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point. In any case the points, that represent these sections, were selected like as shown in code Listing (\ref{code:sectorElevation}). That is a range of angles are defined based on the wind direction d at some distance from the given point. This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.
